---
title: "summarize output"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(sdmTMB)
library(ggplot2)
library(dplyr)
library(viridis)
```

## Total CPUE 

The table below summarizes trends and AIC values for 3 models: a null model (only including depth and depth^2, plus year as a factor), a model with a log-linear trend on epsilon, and a model with a trend on epsilon related to mean temperature. There is some support for the temporal trend model for a number of species, but the only case where the temporal model is supported by AIC (>> 2) is Aurora rockfish. The temporal coefficients are shown in Fig. \ref{fig:trend-cpue}

```{r cpue}
# Species of interest
species = read.csv("survey_data/species_list.csv")
names(species) = tolower(names(species))
species = dplyr::rename(species,
                        common_name = common.name,
                        scientific_name = scientific.name)
species$year_trend = NA
species$year_se = NA
species$temp_trend = NA
species$temp_se = NA
species$aic_null = NA
species$aic_trend = NA
species$aic_trend_temp = NA
species_df = species

for(i in 1:nrow(species_df)){
  comm_name = species_df$common_name[i]
  load(file=paste0("output/", sub(" ", "_", comm_name),"_all_models.RData"))
  indx = which(names(ad_fit_ll$sd_report$value)=="b_epsilon")[1]
  species_df$year_trend[i] = ad_fit_ll$sd_report$value[indx]
  species_df$year_se[i] = ad_fit_ll$sd_report$sd[indx]
  indx = which(names(ad_fit_ll_temp$sd_report$value)=="b_epsilon")[1]
  species_df$temp_trend[i] = ad_fit_ll_temp$sd_report$value[indx]
  species_df$temp_se[i] = ad_fit_ll_temp$sd_report$sd[indx]
  species_df$aic_null[i] = AIC(ad_fit)
  species_df$aic_trend[i] = AIC(ad_fit_ll)
  species_df$aic_trend_temp[i] = AIC(ad_fit_ll_temp)
  
  min_aic <- min(species_df[i,c("aic_null","aic_trend","aic_trend_temp")])
  species_df[i,c("aic_null","aic_trend","aic_trend_temp")] <- species_df[i,c("aic_null","aic_trend","aic_trend_temp")] - min_aic
}

knitr::kable(dplyr::select(species_df,-common_name, -scientific_name,-priority.to.include), digits=2)
```

```{r fig1, fig.cap="Temporal trend in epsilon, for models with total CPUE as a response \\label{fig:trend-cpue}", fig.pos="placeHere"}
p <- species_df %>%
  ggplot(aes(species,year_trend)) +
  geom_pointrange(aes(ymin=year_trend-2*year_se, ymax=year_trend+2*year_se),size=0.8,alpha=0.8,
                  position = position_dodge(width = 0.9)) +
  xlab("Species") +
  ylab("Temporal trend in spatiotemporal sd (+/- 2SE)") +
  geom_hline(aes(yintercept=0),col="red",alpha=0.6) +
  ylim(-0.15,0.15) + 
  coord_flip() +
  theme_bw() +
  scale_color_viridis(discrete=TRUE,end=0.8)
p
```

## Presence-absence 

We ran the same models for presence-absence of each species. 

There is some support for the temporal trend model for a few species (Shortbelly rockfish), but overall support across species seems not strong for the model with epsilon ~ time or model with epsilon ~ temperature. The temporal coefficients are shown in Fig. \ref{fig:trend-presence}

```{r presence}
# Species of interest
species = read.csv("survey_data/species_list.csv")
names(species) = tolower(names(species))
species = dplyr::rename(species,
                        common_name = common.name,
                        scientific_name = scientific.name)
species$year_trend = NA
species$year_se = NA
species$temp_trend = NA
species$temp_se = NA
species$aic_null = NA
species$aic_trend = NA
species$aic_trend_temp = NA
species_df = species

for(i in 1:nrow(species_df)){
  comm_name = species_df$common_name[i]
  load(file=paste0("output/", sub(" ", "_", comm_name),"_presence_models.RData"))
  indx = which(names(ad_fit_ll$sd_report$value)=="b_epsilon")[1]
  species_df$year_trend[i] = ad_fit_ll$sd_report$value[indx]
  species_df$year_se[i] = ad_fit_ll$sd_report$sd[indx]
  indx = which(names(ad_fit_ll_temp$sd_report$value)=="b_epsilon")[1]
  species_df$temp_trend[i] = ad_fit_ll_temp$sd_report$value[indx]
  species_df$temp_se[i] = ad_fit_ll_temp$sd_report$sd[indx]
  species_df$aic_null[i] = AIC(ad_fit)
  species_df$aic_trend[i] = AIC(ad_fit_ll)
  species_df$aic_trend_temp[i] = AIC(ad_fit_ll_temp)
  
  min_aic <- min(species_df[i,c("aic_null","aic_trend","aic_trend_temp")])
  species_df[i,c("aic_null","aic_trend","aic_trend_temp")] <- species_df[i,c("aic_null","aic_trend","aic_trend_temp")] - min_aic
}

knitr::kable(dplyr::select(species_df,-common_name, -scientific_name,-priority.to.include), digits=2)
```

```{r fig2, fig.cap="Temporal trend in epsilon, for models with presence absence as a response \\label{fig:trend-presence}", fig.pos="placeHere"}
p <- species_df %>%
  ggplot(aes(species,year_trend)) +
  geom_pointrange(aes(ymin=year_trend-2*year_se, ymax=year_trend+2*year_se),size=0.8,alpha=0.8,
                  position = position_dodge(width = 0.9)) +
  xlab("Species") +
  ylab("Temporal trend in spatiotemporal sd (+/- 2SE)") +
  geom_hline(aes(yintercept=0),col="red",alpha=0.6) +
  ylim(-0.4,0.4) + 
  coord_flip() +
  theme_bw() +
  scale_color_viridis(discrete=TRUE,end=0.8)
p
```


## Positive models 

We ran the same models for positive catch rates of each species. 

There is some support for the temporal trend model for a few species (Shortbelly rockfish), but overall support across species seems not strong for the model with epsilon ~ time or model with epsilon ~ temperature. The temporal coefficients are shown in Fig. \ref{fig:trend-positive}

```{r positive}
# Species of interest
species = read.csv("survey_data/species_list.csv")
names(species) = tolower(names(species))
species = dplyr::rename(species,
                        common_name = common.name,
                        scientific_name = scientific.name)
species$year_trend = NA
species$year_se = NA
species$temp_trend = NA
species$temp_se = NA
species$aic_null = NA
species$aic_trend = NA
species$aic_trend_temp = NA
species_df = species

for(i in 1:nrow(species_df)){
  comm_name = species_df$common_name[i]
  load(file=paste0("output/", sub(" ", "_", comm_name),"_positive_models.RData"))
  indx = which(names(ad_fit_ll$sd_report$value)=="b_epsilon")[1]
  species_df$year_trend[i] = ad_fit_ll$sd_report$value[indx]
  species_df$year_se[i] = ad_fit_ll$sd_report$sd[indx]
  indx = which(names(ad_fit_ll_temp$sd_report$value)=="b_epsilon")[1]
  species_df$temp_trend[i] = ad_fit_ll_temp$sd_report$value[indx]
  species_df$temp_se[i] = ad_fit_ll_temp$sd_report$sd[indx]
  species_df$aic_null[i] = AIC(ad_fit)
  species_df$aic_trend[i] = AIC(ad_fit_ll)
  species_df$aic_trend_temp[i] = AIC(ad_fit_ll_temp)
  
  min_aic <- min(species_df[i,c("aic_null","aic_trend","aic_trend_temp")])
  species_df[i,c("aic_null","aic_trend","aic_trend_temp")] <- species_df[i,c("aic_null","aic_trend","aic_trend_temp")] - min_aic
}

knitr::kable(dplyr::select(species_df,-common_name, -scientific_name,-priority.to.include), digits=2)
```

```{r fig3, fig.cap="Temporal trend in epsilon, for models with positive catch rate as a response \\label{fig:trend-positive}", fig.pos="placeHere"}
p <- species_df %>%
  ggplot(aes(species,year_trend)) +
  geom_pointrange(aes(ymin=year_trend-2*year_se, ymax=year_trend+2*year_se),size=0.8,alpha=0.8,
                  position = position_dodge(width = 0.9)) +
  xlab("Species") +
  ylab("Temporal trend in spatiotemporal sd (+/- 2SE)") +
  geom_hline(aes(yintercept=0),col="red",alpha=0.6) +
  ylim(-0.4,0.4) + 
  coord_flip() +
  theme_bw() +
  scale_color_viridis(discrete=TRUE,end=0.8)
p
```


## Bottom temperature

We can run the same 3 models used above on the temperature data, with a slight modification in also including a quadratic function of the day of the year (to account for the 2 passes, and duration of the surveys). In this case, there seems to be pretty strong support for modeling epsilon ~ mean temperature.

```{r bottomtemp}
load(file=paste0("output/temp_all_models.RData"))

df = data.frame("model" = c("null","log-linear","mean temp"),
                "AIC"=c(AIC(fit),AIC(fit_ll),AIC(fit_ll_temp)))
df$AIC = df$AIC - min(df$AIC)
knitr::kable(df, digits=2)
```

Digging into the model a bit more, the coefficient is positive (interpretation: years with warmer bottom temperature are also more variable).

```{r}
indx = which(names(fit_ll_temp$sd_report$value)=="b_epsilon")[1]
print(paste("b_epsilon=",round(fit_ll_temp$sd_report$value[indx],2), " SE = ",
            round(fit_ll_temp$sd_report$sd[indx],2)))
```

Fig. \ref{fig:data-temp} shows the deviations by year

```{r fig4, fig.cap="Scaled mean bottom temperature in the WCBTS data, by year \\label{fig:data-temp}", fig.pos="placeHere"}
sub %>% 
  dplyr::group_by(year) %>% 
  summarize(mean_temp = mean_temp[1]) %>% 
  ggplot(aes(year,mean_temp)) + geom_point(size=3) + 
  xlab("Year") + ylab("Standardized mean temperature") + 
  theme_bw()
```
We can also look at the time series of estimated epsilon values by year,

```{r}
indx = which(names(fit_ll_temp$sd_report$value) == "sigma_E")
df = data.frame("Year"=2003:2018,
                sigma = fit_ll_temp$sd_report$value[indx],
                sigma_sd = fit_ll_temp$sd_report$sd[indx])
```


